### 2022-03-14 ehoh/ehoh.tex
past tense

\ourpara{Term Orders} Superposition provers rely on term orders to prune the search
space. The order must be a simplification order that
%can be extended to a simplification order that
% That's true but pointless. It's equivalent to saying that you need a
% simplification order and the prover may underapproximate it. -JB
is total on variable-free
terms. E implements both the Knuth--Bendix order (KBO) and the lexicographic path order
(LPO). KBO is widely regarded as the more
robust option for superposition.
%
In earlier work, Blanchette and colleagues have shown that only KBO can be
generalized gracefully while preserving the necessary properties for
superposition \cite{bbww-17-kbo,bww-17-rpo}. For this
reason, we focus on KBO.

@@

There is another transformation performed by preprocessing that is
problematic, but for a different reason. \emph{Definition unfolding} is the
process of replacing equationally defined symbols with their definitions and
removing the defining equations. A \relax{definition} is a clause of the form
$\cst{f} \; \tuple{x}{m} \eq t$, where the variables $\tuple{x}{m}$ are distinct,
$\cst{f}$ does not occur in the right-hand side $t$, and $\Var(t) \subseteq
\{x_1,\dots,x_m\}$.
This transformation preserves unsatisfiability (provability) for first-~and
higher-order logic, but not for \lfhol{}, making Ehoh incomplete. The reason
is that by removing the definitional clause, we also remove a symbol $\cst{f}$
that otherwise could be used to instantiate a higher-order quantifier.
For example, the clause set
$\{\cst{f}\;x \eq x{,}\; \cst{f}\;(y\;\cst{a}) \noteq \cst{a}\}$ is unsatisfiable,
whereas $\{y\;\cst{a} \noteq \cst{a}\}$ is satisfiable in \lfhol{}. (In full
higher-order logic, the second clause set would be unsatisfiable thanks to
the $\{y \mapsto \lambda x.\; x\}$ instance and $\beta$-conversion.)
%
For the moment, we have simply disabled definition unfolding in Ehoh. We have
not measured the effect of this choice, but we conjecture it is not substantial.
Namely, appropriately chosen term order will result in a similar outcome without
removing the defined symbol. We will enable definition unfolding again once we
have added support for $\lambda$-terms.

>>>>

\ourpara{Term Orders} Superposition provers rely on term orders to prune the search
space. The order must be a reduction order
%can be extended to a simplification order that
% That's true but pointless. It's equivalent to saying that you need a
% simplification order and the prover may underapproximate it. -JB
(Sect.~\ref{sec:pre:order}). E implements both the Knuth--Bendix order (KBO) and the lexicographic path order
(LPO). KBO is widely regarded as the more
robust option for superposition.
%
In earlier work, Blanchette and colleagues have shown that only KBO can be
generalized gracefully while preserving the necessary properties for
superposition \cite{bbww-17-kbo,bww-17-rpo}. For this
reason, we focus on KBO.

@@

There is another transformation performed by preprocessing that is
problematic, but for a different reason. \emph{Definition unfolding} is the
process of replacing equationally defined symbols with their definitions and
removing the defining equations. A \relax{definition} is a clause of the form
$\cst{f} \; \tuple{x}{m} \eq t$, where the variables $\tuple{x}{m}$ are distinct,
$\cst{f}$ does not occur in the right-hand side $t$, and $\Var(t) \subseteq
\{x_1,\dots,x_m\}$.
This transformation preserves unsatisfiability (provability) for first-~and
higher-order logic, but not for \lfhol{}, making Ehoh incomplete. The reason
is that by removing the definitional clause, we also remove a symbol $\cst{f}$
that otherwise could be used to instantiate a higher-order quantifier.
For example, the clause set
$\{\cst{f}\;x \eq x{,}\; \cst{f}\;(y\;\cst{a}) \noteq \cst{a}\}$ is unsatisfiable,
whereas $\{y\;\cst{a} \noteq \cst{a}\}$ is satisfiable in \lfhol{}. (In full
higher-order logic, the second clause set would be unsatisfiable thanks to
the $\{y \mapsto \lambda x.\; x\}$ instance and $\beta$-reduction.)
%
For the moment, we have simply disabled definition unfolding in Ehoh. We have
not measured the effect of this choice, but we conjecture it is not substantial.
Namely, appropriately chosen term order will result in a similar outcome without
removing the defined symbol. We will enable definition unfolding again once we
have added support for $\lambda$-terms.

### 2022-03-14 introduction/introduction.tex
past tense

These attempts resulted in early higher-order theorem provers. Andrews and
colleagues developed TPS~\cite{abinpx-96-tps}: an automatic prover which
allows users to specify proof outlines, based on expansion proofs. Benzm\"uller
and colleagues developed LEO \cite{cbmk-98-leo}, a prover based on higher-order
resolution which introduced the cooperative paradigm. Provers implementing this paradigm
invoke first-order backends in regular intervals to finish the proof. The latest
iteration in the LEO family of provers is Leo-III \cite{sb-21-leo3}. Satallax
\cite{cb-12-satallax} is based on higher-order analytic tableau and uses SAT solver
to close tableau branches. It won the higher-order division of CASC on eight occasions:
in 2011 and every year from 2013 to 2019.

>>>>

\looseness=-1
These attempts resulted in early higher-order theorem provers. Andrews and
colleagues developed TPS~\cite{abinpx-96-tps}: an automatic prover which allows
users to specify proof outlines, based on expansion proofs. Benzm\"uller and
colleagues developed LEO \cite{cbmk-98-leo}, a prover based on higher-order
resolution which introduced the cooperative paradigm. Provers implementing this
paradigm invoke first-order backends in regular intervals to finish the proof.
The latest iteration in the LEO family of provers is Leo-III \cite{sb-21-leo3}.
Satallax \cite{cb-12-satallax} is based on higher-order analytic tableaux. Such
representations of logical formulas are tree structures with at each node a
subformula of the original formula to be proved or refuted. Satallax improves on
the method by using a SAT solver to close tableau branches.
It won the higher-order division of CASC on eight occasions: in 2011 and every
year from 2013 to 2019.

### 2022-03-14 pre/pre.tex
past tense

\looseness=-1
With $\alpha$-renaming we assume the rule $(\lambda x. \, s) \longrightarrow \lambda
y. \, \{x \mapsto y\}(s)$ where $y$ does not occur loosely bound in $s$ and 
$y$ is not bound in $s$. Application of substitution $\sigma$ to term
$t$, implicitly $\alpha$-renames bound variables to avoid capture.  For example,
$\{X \mapsto x\}(\lamx{X \, \cst{a}})$ results in $\lam{y}{x \, \cst{a}}$. The
$\beta$-reduction rule corresponds to applying arguments to function: $(\lambda
x. s) \, t \longrightarrow \{x \mapsto t\} (s)$, where the bound variables in $s$
are renamed to avoid capture. Lastly, $\eta$-reduction is defined as $\lambda
x.\, s \, x \longrightarrow s$ under the condition that $x$ is not loosely
bound in $x$. We write $s \equi t$ if terms $s$ and $t$ are equal modulo $\alpha\beta\eta$
converison.

@@

\looseness=-1
An {\em equation} $s \eq t$ corresponds to an unordered pair of terms. A {\em
literal} $l$ is an equation $s \eq t$ or its negation (disequation) $s \noteq
t$. A clause $C$ is a finite multiset of literals, interpreted and written
disjunctively:\ $l_1 \llor \cdots \llor l_n$. Free variables of clauses are
implicitly universally quantified. Note that $\llor$ denoting disjunction in
clauses is different from the formula-building $\ior$. In general,
formula-building logical connectives are written in boldface.

In standard, non-clausal first-order logic, an atom is either
predicate or equational. For uniformity, and to
stay close to the implementation, we encode predicate atoms as equations with
$\itrue$. Negative literals are encoded as disequations. For example,
$\cst{even}(x)$ is encoded as $\poslit{\cst{even}(x)}$, and
$\neg\,\cst{even}(x)$ is encoded as $\neglit{\cst{even}(x)}$. To lighten the notation,
we sometimes write the predicate literals in non-encoded form.

@@

Superposition is one of the most successful calculi for first-order logic with
equality. It owes its success to a careful treatment of equality and built-in
heuristics to prune the search space such as term order and selection functions.
Before introducing the rules of the calculus, we provide definitions of these and other
background concepts.

@@

In higher-order logic, unification is performed modulo rules of
$\alpha\beta\eta$ conversion. Under these conditions, the uniqueness of the most
general unifier is not guaranteed. Consider the constraint $X \,
(\cst{f} \, \cst{a}) \unif \cst{f} \, (X \, \cst{a})$. Any substitution of the
form $\{ X \mapsto \lamx{\cst{f}^i \, x} \}$ is a unifier, with $\cst{f}^i \, x$
denoting iterated, $i$-fold application of $\cst{f}$. However, neither of these substitutions
is more general than the other. To generalize most general
unifiers to higher-order logic, the concept of the {\em complete sets of unifiers} (CSU) was
introduced.

@@

Lastly, the rules \infname{SP} and \infname{SN} are used to simulate what mathematicians do when
they ``replace equals by equals''. The left premise intuitively states that an
equation $s \eq t$ holds under condition $C$ (or more precisely, its negation).
Similarly, the right premise asserts that a (dis)equation holds under a
condition. The superposition rule simply concatenates the conditions for both
premises and replaces equals by equals in two main inference terms. For example
given clauses $\neg \cst{even}(X) \llor \cst{f}(X) \eq \cst{a}$ and $\neg
\cst{even}(\cst{b}) \llor \cst{g}(\cst{f}(Y)) \eq \cst{b}$, superposition between their
last literals results in $\neg
\cst{even}(X) \llor \neg \cst{even}(\cst{b}) \llor \cst{g}(\cst{a}) \eq
\cst{b}$.

>>>>

\looseness=-1
With $\alpha$-conversion we assume the rule $(\lambda x. \, s) \longrightarrow
\lambda y. \, \{x \mapsto y\}(s)$ where $y$ does not occur loosely bound in $s$
and $y$ is not bound in $s$. Application of a substitution to a term,
implicitly $\alpha$-renames bound variables to avoid capture.  For example, $\{X
\mapsto x\}(\lamx{X \, \cst{a}})$ results in $\lam{y}{x \, \cst{a}}$. The
$\beta$-reduction rule, roughly speaking, corresponds to passing arguments in a
function call: $(\lambda x. s) \, t \longrightarrow \{x \mapsto t\} (s)$, where
the bound variables in $s$ are renamed to avoid capture. Lastly,
$\eta$-reduction is defined as $\lambda x.\, s \, x \longrightarrow s$ under the
condition that $x$ is not loosely bound in $x$. We write $s \equi t$ if terms
$s$ and $t$ are equal modulo $\alpha\beta\eta$-conversion. Terms that are
convertible with respect to any of the rules form equivalence classes. Thus, we
also say that $\alpha$($\beta\eta$)-convertible terms are
$\alpha$($\beta\eta$)-equivalent.

@@

\looseness=-1
An {\em equation} $s \eq t$ corresponds to an unordered pair of terms. A {\em
literal} $l$ is an equation $s \eq t$ or its negation (disequation) $s \noteq
t$. A clause $C$ is a finite multiset of literals, interpreted and written
disjunctively:\ $l_1 \llor \cdots \llor l_n$. Free variables of clauses are
implicitly universally quantified. Note that $\llor$ denoting disjunction in
clauses is different from the formula-building $\ior$. In general,
formula-building logical connectives are written in boldface.
In standard, non-clausal first-order logic, an atom is either
predicate or equational. For uniformity, and to
stay close to the implementation, we encode predicate atoms as equations with
$\itrue$. Negative literals are encoded as disequations. For example,
$\cst{even}(x)$ is encoded as $\poslit{\cst{even}(x)}$, and
$\neg\,\cst{even}(x)$ is encoded as $\neglit{\cst{even}(x)}$. To lighten the notation,
we sometimes write the predicate literals in non-encoded form.

@@

Superposition is one of the most successful calculi for first-order logic with
equality. It owes its success to a careful treatment of equality and built-in
heuristics to prune the search space such as term order and selection functions.

This calculus works on problems in clausified form and proves the problem by
refuting its negation. In practice this means that, to apply superposition to a
given problem, we first must negate its conjecture and clausify the whole
problem using some of the clausification algorithms
\cite{nw-01-small-cnf,rsv-16-vcnf}. The calculus then applies its inference
rules to the clauses, adding results of the inferences to the working set of
clauses. As this process is quite prolific it also uses the term order to determine
if some clauses can be simplified or eliminated from the working set.
Superposition is refutationally complete: This means that, if results of all
inferences are computed in a systematic, fair manner, empty clause ($\bot$) will
eventually be derived for a provable problem. As first-order logic is
undecidable, no guarantees are given for unprovable problems.
Before introducing the rules of the calculus, we provide definitions of these and other
background concepts.

@@

In higher-order logic, unification is performed modulo rules of
$\alpha\beta\eta$-conversion. Under these conditions, the uniqueness of the most
general unifier is not guaranteed. Consider the constraint $X \,
(\cst{f} \, \cst{a}) \unif \cst{f} \, (X \, \cst{a})$. Any substitution of the
form $\{ X \mapsto \lamx{\cst{f}^i \, x} \}$ is a unifier, with $\cst{f}^i \, x$
denoting iterated, $i$-fold application of $\cst{f}$. However, neither of these substitutions
is more general than the other. To generalize most general
unifiers to higher-order logic, the concept of the {\em complete sets of unifiers} (CSU) was
introduced.

@@

Lastly, the rules \infname{SP} and \infname{SN} are used to simulate what
mathematicians do when they ``replace equals by equals''. Name paramodulation is
also used for rules that replace equals by equals in calculi preceding superposition.  The left premise intuitively
states that an equation $s \eq t$ holds under condition $C$ (or more precisely,
its negation). Similarly, the right premise asserts that a (dis)equation holds
under a condition. The superposition rule simply concatenates the conditions for
both premises and replaces equals by equals in two main inference terms. For
example given clauses $\neg \cst{even}(X) \llor \cst{f}(X) \eq \cst{a}$ and
$\neg \cst{even}(\cst{b}) \llor \cst{g}(\cst{f}(Y)) \eq \cst{b}$, superposition
between their last literals results in $\neg \cst{even}(X) \llor \neg
\cst{even}(\cst{b}) \llor \cst{g}(\cst{a}) \eq \cst{b}$.
